Road Segment Prioritization for Bicycle Infrastructure

### 2020-11-13 README.Rmd
add pdf
add pdf and rmd
paper intro done

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography: mybibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that is successful in getting more people to adopt cycling as a mode of transport. Therefore, the aim of this study is to develop a methodology that determines where cycling infrastructure should be added by accounting for cyclist preference and UK government policies to create Low Traffic Neighborhoods. Distance decay, routing, and network analysis techniques are used to examine where future cycling demand in a city will come from, how such demand should be routed on the street network, and how to  consider fair distribution of cycling infrastructure that is in line with egalitarian principles. The city of Manchester is used as a case study, but the methodology is generic enough to be applicable to any city. For routing, it is found that deviating from shortest paths to avoid high-stress road segments can result in acceptable travel time increases for cyclists. Community detection is used to partition the city and evaluate results from utilitarian to egalitarian algorithms on distributing cycling infrastructure. The results show that the egalitarian algorithm provides comparable city-wide results to the utilitarian one. \\par\\textbf{Keywords:} cycling network, routing, transport equity"
# keywords: "cycling, potential demand, routing"
output:
  # bookdown::github_document2
  bookdown::pdf_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable"]
---

@@

- Lit Review: I can also organize this thematically and trim down **Planning Cycling Networks** substantially. I would then add some literature in:
  - Calculating Potential Demand
  - Routing
  - Road Segment Prioritization
- Community Detection: Where in the document should this section be?
- Other Cities

@@

Intro ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

# Background

## What Affects the Decision To Cycle

Segregated cycling infrastructure^[Segregated cycling infrastructure refers to road space that is allocated to cyclists only, with physical separation to protect cyclists from other modes of transport.] has been shown to increase cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators]. Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal]. However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far]. Another defining feature for cycling infrastructure is how well connected it is. Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter], and the lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining]. While direct and cohesive cycling networks have been shown to positively impact cycling rates, density^[making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options] of the cycling network is also vital [@schoner2014missing]].

## Planning Cycling Networks

*Optimization* techniques have been used to propose improvements to cycling networks. @mesbah2012bilevel propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget. The upper level is the proposed interventions and the lower level is the route choices made by users in reaction to changes in the network. The problem accounts for the effect of cycling lanes on car traffic, and attempts to maximize utilization of said lanes with minimal impact on car travel times. To improve cohesion of the suggested network, a constraint is added so that each link with a bike lane should be connected to at least one destination. Car usage is not considered by @mauttone2017bicycle, who develop an optimization framework that aims to minimize the total user cost of cycling on the network. The aggregate flow on links is obtained by using shortest paths to route existing cycling demand onto the road network, and the solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network. The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure, and a discontinuity penalty is also added to prioritize connected road segments. The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization]. In this formulation, all OD pairs need to be connected by roads that meet the desired LOS, and a directness constraint is added so that paths between OD pairs do not exceed a certain multiple of the shortest path.

\vspace{0.3cm}

These problem formulations do not explicitly solve for continuity, which is dealt with using a either (a) a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], (b) a constraint on deviation from shortest paths [@duthie2014optimization], or (c) a discontinuity penalty [@mauttone2017bicycle]. To solve for continuity, the graph-theoretic concept of *connected components*, has been used. @natera2019data study the existing cycling network in terms of its disconnected components and introduce two different algorithms to connect these components by their most critical links^[*link* refers to a road segment throughout this research] and, in doing so, measure the size of the growth of the largest connected component as a function of the kilometers of network added. They observe that small investments at strategic points have a large impact on connectivity in most cases. The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow^[*flow* is used throughout this research to refer to the cycling demand when it is routed onto the road network. The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various OD pairs] passing through them. They vary the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a giant component. The results show a cycling network that connects the entire city, and subtracting links intersecting with current cycling infrastructure identifies links proposed for intervention.

\vspace{0.3cm}

@@

## Underlying Ethical Principles

The methodologies in Section \@ref(planning-cycling-networks) are underpinned by different ethical principles, even though these principles are not explicitly acknowledged by the authors. This is important since different ethical principles constitute different problem formulations and targets. Broadly speaking, transport appraisal can be based on either utilitarian or egalitarian principles. The former seeks to maximize the overall benefit, while the latter is concerned with a fair distribution of benefits [@jafino2020transport]. @nahmias2017integrating criticize the utilitarian approach that has been historically popular in the evaluation of transport investments, explaining how the maximization of overall benefit fails to account for the distribution of that benefit among communities or individuals. @lucas2016method explain how transport studies have traditionally looked at the bigger picture without studying the distribution of investments on the different parts of the study area, and go on to propose an egalitarian approach that ensures the dis-aggregation of transport policy benefits across the study area. @pereira2017distributive also emphasize the need for a more egalitarian approach to transport planning. They highlight accessibility as a cornerstone of distributive justice, and contend that policies should aim to distribute investments in a way that minimizes spatial variations in accessibility. This research attempts to provide a methodology that is grounded in egalitarian principles. **Write some more here**

@@

The analysis is heavily dependant on Origin-Destination census data (commuter data). Commuter data in the UK is publicly available at the Middle layer Super Output Area (MSOA) level; the average MSOA has a population of 8209 [@ofn2018population]. @iacono2010measuring note that such large travel zones are not ideal for understanding route choice behaviour of cyclists and pedestrians. They also give rise to an \`ecological fallacy' whereby average characteristics are assumed to apply to all residents of the aggregated geographical area. Given that more granular data is not publicly available, the study uses MSOA-level commuter data. The methodology is however applicable to more granular commuter data should it become available.

@@

# Conclusions

>>>>

---
title: Road Segment Prioritization for Bicycle Infrastructure
date: "`r Sys.Date()`"
geometry: margin=2cm
bibliography: mybibfile.bib
#linenumbers: true
numbersections: true
abstract: "Understanding the motivators and deterrents to cycling is essential for creating infrastructure that is successful in getting more people to adopt cycling as a mode of transport. Therefore, the aim of this study is to develop a methodology that determines where cycling infrastructure should be added by accounting for cyclist preference and UK government policies to create Low Traffic Neighborhoods. Distance decay, routing, and network analysis techniques are used to examine where future cycling demand in a city will come from, how such demand should be routed on the street network, and how to  consider fair distribution of cycling infrastructure that is in line with egalitarian principles. The city of Manchester is used as a case study, but the methodology is generic enough to be applicable to any city. For routing, it is found that deviating from shortest paths to avoid high-stress road segments can result in acceptable travel time increases for cyclists. Community detection is used to partition the city and evaluate results from utilitarian to egalitarian algorithms on distributing cycling infrastructure. The results show that the egalitarian algorithm provides comparable city-wide results to the utilitarian one. \\par\\textbf{Keywords:} cycling network, routing, transport equity"
# keywords: "cycling, low-traffic neighborhoods, potential demand, routing"
output:
  # bookdown::github_document2
  bookdown::pdf_document2:
      toc: false
      extra_dependencies: ["subcaption", "longtable"]
---

@@

- Lit Review: I can also organize this thematically and trim down **Planning Cycling Networks** substantially. I would then add some literature in:
  - Calculating Potential Demand
  - Routing
  - Road Segment Prioritization
- Community Detection: Where in the document should this section be?
- Should we use other cities for the sake of comparison
- How to add tables
- Footnotes are sometimes split between pages. How do I keep them on the same page?

@@

The 2015 Paris agreement [@agreement2015paris] was an acknowledgment that fundamental changes to our lives and economies are necessary to reduce the impacts of climate change. The transport sector is one of many sectors facing transformation, with innovation and policy measures leading to the introduction of ‘greener’ private vehicles. But as demand for transport rises, resulting in increased congestion, public and active modes of transport are gaining more recognition. The benefits of the latter extend beyond congestion and the environment, as it promises to help alleviate what is referred to as the pandemic of global inactivity; Physical inactivity is on the rise and has become the 4th highest cause of death globally [@kohl2012pandemic]. Various studies have documented the association between active transport and lower risk of disease, including cancer and cardiovascular disease [@celis2017association; @jarrett2012effect; @patterson2020associations].
In the wake of the Covid-19 pandemic, and the resulting reduced capacity of public transport, the UK government has pledged to invest billions of pounds to improve walking and cycling infrastructure across the country. While this unprecedented sum is an opportunity to reshape cities in a way that improves the well-being of citizens, it does come with a warning:

> "Inadequate cycling infrastructure discourages cycling and wastes public money. Much cycling infrastructure in this country is inadequate. It reflects a belief, con- scious or otherwise, that hardly anyone cycles, that cycling is unimportant and that cycles must take no meaningful space from more important road users, such as motor vehicles and pedestrians - [@departmentgearchange2020]"

The funding on its own is therefore no guarantee of a change in commuting across the country; it must be used to design adequate cycling infrastructure that is based on motivators and deterrents to cycling. A considerable amount of research has been done towards that end. Segregated cycling infrastructure^[Segregated cycling infrastructure refers to road space that is allocated to cyclists only, with physical separation to protect cyclists from other modes of transport.] has been shown to increase cycling uptake [@aldred2019impacts; @goodman2014new; @marques2015infrastructure], with the separation from motorized vehicles being key [@winters2011motivators]. Revealed preference of cyclists shows that they are willing to deviate from the most efficient routes in order to commute on safer roads [@crane2017longitudinal]. However, such deviations are only considered if they do not considerably increase route circuitry; behaviour studies have found that the probability of choosing a route decreases in proportion to its length relative to the shortest route [@broach2011bicycle; @winters2010far]. Another defining feature for cycling infrastructure is how well connected it is. Cyclists prefer cohesive infrastructure, particularly when cycling on arterial roads with high levels of motorized traffic [@stinson2003commuter], and the lack of well-connected cycling infrastructure is one of the main obstacles to increasing cycling uptake [@caulfield2012determining]. While direct and cohesive cycling networks have been shown to positively impact cycling rates, density^[making an area's bicycle network denser means adding more cycling routes in the area and thereby giving cyclists more route options] of the cycling network is also vital [@schoner2014missing].

As noted by @buehler2016bikeway, it is only recently that research has pivoted from focusing on localized, street or intersection level impact towards studying the whole cycling network. They emphasize the promise of this shift in capturing the network-wide effect of street-level interventions. *Optimization* techniques have been used to propose improvements to cycling networks. @mesbah2012bilevel propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget. The upper level is the proposed interventions and the lower level is the route choices made by users in reaction to changes in the network. The problem accounts for the effect of cycling lanes on car traffic, and attempts to maximize utilization of said lanes with minimal impact on car travel times. To improve cohesion of the suggested network, a constraint is added so that each link with a bike lane should be connected to at least one destination. Car usage is not considered by @mauttone2017bicycle, who develop an optimization framework that aims to minimize the total user cost of cycling on the network. The aggregate flow on links is obtained by using shortest paths to route existing cycling demand onto the road network, and the solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network. The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure, and a discontinuity penalty is also added to prioritize connected road segments. The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization]. In this formulation, all OD pairs need to be connected by roads that meet the desired LOS, and a directness constraint is added so that paths between OD pairs do not exceed a certain multiple of the shortest path.

These problem formulations do not explicitly solve for continuity, which is dealt with using a either (a) a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], (b) a constraint on deviation from shortest paths [@duthie2014optimization], or (c) a discontinuity penalty [@mauttone2017bicycle]. To solve for continuity, the graph-theoretic concept of *connected components*, has been used. @natera2019data study the existing cycling network in terms of its disconnected components and introduce two different algorithms to connect these components by their most critical links^[*link* refers to a road segment throughout this research] and, in doing so, measure the size of the growth of the largest connected component as a function of the kilometers of network added. They observe that small investments at strategic points have a large impact on connectivity in most cases. The concept of connected components is also at the core of the methodology proposed by @olmos2020data. After routing the cycling demand onto the network links, they use percolation theory to filter out the links based on the aggregate flow^[*flow* is used throughout this research to refer to the cycling demand when it is routed onto the road network. The flow on any road segment is the cumulative demand on it, resulting from cyclists commuting between various OD pairs] passing through them, varying the flow threshold for filtering to identify the minimum flow at which the whole city is connected by a giant component. The results show a cycling network that connects the entire city, and subtracting links intersecting with current cycling infrastructure identifies links proposed for intervention.

@@

All of these network-level methodologies are underpinned by different ethical principles, even though these principles are not explicitly acknowledged by the authors. This is important since different ethical principles constitute different problem formulations and targets. Broadly speaking, transport appraisal can be based on either utilitarian or egalitarian principles. The former seeks to maximize the overall benefit, while the latter is concerned with a fair distribution of benefits [@jafino2020transport]. @nahmias2017integrating criticize the utilitarian approach that has been historically popular in the evaluation of transport investments, explaining how the maximization of overall benefit fails to account for the distribution of that benefit among communities or individuals. @lucas2016method explain how transport studies have traditionally looked at the bigger picture without studying the distribution of investments on the different parts of the study area, and go on to propose an egalitarian approach that ensures the dis-aggregation of transport policy benefits across the study area. @pereira2017distributive also emphasize the need for a more egalitarian approach to transport planning. They highlight accessibility as a cornerstone of distributive justice, and contend that policies should aim to distribute investments in a way that minimizes spatial variations in accessibility. This research attempts to compare the two principles, and in doing so determine whether a methodology formulated based on an egalitarian approach can be feasible in designing a cycling network that aligns with motivators and deterrents to cycling. **Write some more here**

@@

The analysis is heavily dependent on commuter data^[*commuter data* refers to publicly available Origin-Destination (OD) data, normally obtained from a census. This study uses data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between MSOA pairs, by mode of travel.]. Commuter data in the UK is publicly available at the Middle layer Super Output Area (MSOA) level; the average MSOA has a population of 8209 [@ofn2018population]. @iacono2010measuring note that such large travel zones are not ideal for understanding route choice behaviour of cyclists and pedestrians. They also give rise to an \`ecological fallacy' whereby average characteristics are assumed to apply to all residents of the aggregated geographical area. Given that more granular data is not publicly available, the study uses MSOA-level commuter data. The methodology is however applicable to more granular commuter data should it become available.

@@

<!-- # Conclusions -->
